---
title: "Project Part 1"
author: "Drashti Patel (7964120)"
date: "11/01/2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
fifa <- read.csv("FIFA.csv")
```

Here's a description of my data set.
# This is a dataset of performance statistics for elite FIFA players during the 2024 season, including metrics such as ratings(overall), potential,age, and Hits.

Here's the URL where I found my dataset.
# https://www.kaggle.com/datasets/aayushmishra1512/fifa-2021-complete-player-data

My response variable is Overall.

My two explanatory variables are Hits and Potential.

```{r}
fifa1 <- lm(Overall ~ Hits, data = fifa)
summary(fifa1)

# R^2 = 0.4385, Adj_r^2 = 0.4268

fifa2 <- lm(Overall ~ Potential, data = fifa)
summary(fifa2)

# R^2 = 0.5808, Adj_r^2 = 0.572
```

For $Y\sim X1$, $r^2=0.4385$ and $R^2_{adj}=0.4268$
For $Y\sim X2$ $r^2=0.5808$ and $R^2_{adj}=0.572$

```{r}
fifa_combined <- lm(Overall ~ Potential+ Hits, data = fifa)
summary(fifa_combined)

# Adjusted R^2 = 0.6103
```
For $Y\sim X1 + X2$ $R^2_{adj}=0.6103$